---
title: "Statistical report: Chronic obstructive pulmonary disease (COPD) in SwedeHF"
subtitle: DRAFT
date: "`r Sys.Date()`"
cover: static/ki_logo_vert_rgb.png
editor: source
format:
  pdf:
    documentclass: scrartcl
    template-partials:
      - "static/before-body.tex"
      - "static/_titlepage.tex"
    include-in-header: 
      - "static/in-header.tex"
    toc: true
    toc-depth: 3
    toc-title: Contents
    lof: true
    lot: true
    cap-location: top
    number-sections: true
    colorlinks: false
    keep-tex: false
    
bibliography: static/references.bib
csl: static/bmj.csl
link-citations: true
link-bibliography: true
nocite: '@*'
knitr:
  opts_chunk: 
    dev: cairo_pdf
    collapse: true
    comment: "" 
    echo: FALSE
    include: TRUE
    warning: FALSE
    message: FALSE
    fig.pos: "H"
    fig.path: "../output/figs/"
    R.options:
      knitr.graphics.auto_pdf: true
---

\clearpage

```{r}
#| label: set-up-load-data
options(knitr.kable.NA = "")

# load packages, globals and project specific functions
source(here::here("setup/setup.R"))

# load data
load(here("data/clean-data/data.RData"))

# create workbook to write tables to Excel
wb <- createWorkbook()
```           

# Data

## Data sources

The Swedish Heart Failure Registry (SwedeHF) linked with national Swedish registries (SHFDB v 4.1.0) through the Swedish PIN number[@pin]. 

SwedeHF is a nationwide health quality and research registry started in 2000, 
that includes in- and out-hospital patients with HF, regardless of EF and previous history of HF[@swedehf]. 
Around 80 variables are recorded at 
discharge from hospital or at the outpatient visit, i.e. index date, and 
entered into an electronic database managed by the Uppsala Clinical Research 
Center (Uppsala, Sweden). Up to April 2017 the only inclusion criterion was a 
clinical diagnosis of HF, which was thereafter defined according to the ICD-10 
codes I50.0, I50.1, I50.9, I42.0, I42.6, I42.7, I25.5, I11.0, I13.0, I13.2. 
Informed consent is not required, although patients are informed of entry and can opt out. 
In 2021, 69 of 76 Swedish hospitals enrolled patients in SwedeHF, which had a 32% coverage of the prevalent HF in Sweden[@annualreport]. 

- National Patient Register (The National Board of Health and Welfare) for additional comorbidities and the hospitalization outcomes
- Cause of Death Register (The National Board of Health and Welfare) for outcomes death
- National Dispensed Drug Register (The National Board of Health and Welfare) for COPD medications
- LISA[@lisa] and Register of the Total Population (Statistics Sweden) for socio-economic factors

## Ethics

The study was approved by the Swedish Ethical Review Authority, dnr 2021-04326. 

## Acknowledgments

Please include "We thank all staff members at all care units in Sweden for their contribution to the SwedeHF register." 
in the Acknowledgements of any resulting articles. 

## Inclusion/exclusion criteria

```{r}
#| label: flow
#| tbl-cap: Information for flowchart
default_kable(flow) %>%
  row_spec(c(1, 10), bold = T)
```

## Definitions

The definitions for the general dataset, including ICD-10 codes used for outcomes and co-morbidities, are found https://kiheartfailure.github.io/shfdb4/definitions/.

Additional variables created for this project: 

```{r}
#| label: npr
#| tbl-cap: Additional comorbidites from the National Patient Register
default_kable(metaout)
```

```{r pdr}
#| label: pdr
#| tbl-cap: Additional medications from the National Prescribed Drug Register
default_kable(metalm) %>%
  column_spec(2, width = "15cm")
```

# Statistical analysis 

## General

All analyses were performed using `r sessionInfo()$R.version$version.string`[@r]. 
The level of significance is set to 5%, two-sided. No adjustment for multiple 
comparisons were made and therefore the results should be viewed with some care.

## Missing data

Missing data was imputed with multiple imputation (n = 10) using mice[@mice]. 
Variables included in the model are indicated in 
@tbl-base. The primary outcome, all-cause mortality/first HF hospitalization, was included as the Nelson-Aalen estimator.

## COPD

Presentations will be by: 

1) No COPD/COPD.

2) No COPD/COPD without COPD-associated hospitalisations in the previous 12 months/COPD with $\ge$ 1 COPD-associated hospitalisations in the previous 12 months. 

Where the 2 group presentation is considered to be primary. 

## EF subgroups 

Presentations will, were applicable, be by EF subgroups. 

## Baseline characteristics

Baseline characteristics are presented by COPD 
were categorical variables are presented with n and percentage and tested between groups with the chi-square test. 
Continuous variables are presented with median (q1-q3) and tested for differences between groups with the Kruskal-Wallis test. 

## Association between predictors and COPD

The association between predictors and COPD was evaluated using a multivariable logistic regression 
including the variables shown in @fig-predictors. 

Subgroup analyses within each EF was performed including an interaction between EF and the respective variable. 
Note that this causes the estimates to be from different multivariable models.  

### Assumptions

Possible outliers were evaluated using Cook's distance and multicollinearity was 
evaluated using the Variance Inflation Factor (VIF). 

## Association between COPD and outcomes

The following outcomes are considered: 

- CV mortality + first HF hospitalization (primary endpoint)
- CV mortality
- First HF hospitalization
- Total HF hospitalization
- All-cause mortality
- Non-CV mortality

Time to first events were presented with cumulative incidence curves for the primary endpoint. 

Cox proportional hazards regressions were used to model the time to first event. 

As a consistency analysis the time to first events for the primary outcome was
modelled using a sub-distributional hazards model[@fg] 
where death (other than CVD) was treated as a competing event. 

Recurrent events were model using a negative binomial regression including the log of time as an offset in the model. 

Data were censored at 5 years follow-up, 2021-12-31, emigration from Sweden or, when it was itself not an event, death. 

Crude and adjusted analyses were performed were adjustment was for variables indicated in @tbl-base. 

### Association between COPD and outcomes - Subgroups

Associations across pre-specified subgroups were evaluated using an interaction term between COPD and the 
respective variable for the outcome All-cause mortality/first HF hospitalization (primary outcome) and presented in a Forest plot. 

### Assumptions

The proportional hazards assumption was investigated using the scaled Schoenfeld 
residuals (cox.zph in[@survival-package]) for the primary outcome leading to 
that location and age were stratified for in the models. 
Possible outliers were visually inspected by plotting the dfbetas. 

# Results

```{r}
med <- rsdata %>%
  summarise(
    med = fn(median(shf_age), dig = 1),
    q1 = fn(quantile(shf_age, probs = 0.25), dig = 1),
    q3 = fn(quantile(shf_age, probs = 0.75), dig = 1)
  ) %>%
  mutate(out = paste0(med, " (", q1, "-", q3, ")")) %>%
  pull(out)

percfem <- rsdata %>%
  count(shf_sex) %>%
  mutate(perc = fn(n / sum(n) * 100, 1)) %>%
  filter(shf_sex == "Female") %>%
  pull(perc)
```

The median age (IQR) is `r med` and `r percfem`% females.    

## Baseline characteristics

{{< include src/base_table.qmd >}}

## Association between predictors and COPD

{{< include src/predictors.qmd >}}

## Association between COPD and outcomes

```{r}
medsurv <- rsdata %>%
  summarise(
    med = fn(median(sos_outtime_death / 365.25 * 12), dig = 1),
    min = fn(min(sos_outtime_death / 365.25 * 12), dig = 1),
    max = fn(max(sos_outtime_death / 365.25 * 12), dig = 1)
  ) %>%
  mutate(out = paste0(med, " (", min, "-", max, ")")) %>%
  pull(out)

sumsurv <- rsdata %>%
  summarise(sumpy = fn(sum(sos_outtime_death) / 365.25, dig = 0)) %>%
  pull(sumpy)
```

The median (min-max) follow-up is 
`r medsurv` months for a total of `r sumsurv` patient-years of follow-up.

{{< include src/outcomes.qmd >}}

{{< include src/outcomes_subgroups.qmd >}}

{{< include src/km_figs.qmd >}}

\clearpage
\newpage

# Reproducibility

## R session information {#sessioninfo}

```{r sessinfo}
sessionInfo()
```

## R code

The R code for all data handling and statistical analyses are found: 

https://github.com/KIHeartFailure/swedehf-copd. On publication
the repository will be made public so as to 
link to it from the resulting article for increased transparency and code sharing.
No data or output is stored in the repository. 

\clearpage
\newpage

# References

::: {#refs}
:::

